Ligeti, Stockhausen and Koenig:\\Aesthetics of Electronic Music

### 2015-09-27 manuscript.tex
Fix quote overflow on right margin
Apply TODOs

\mdfdefinestyle{MyShadeQuoteStyle}{%
    leftmargin=15pt,
    rightmargin=15pt,
    backgroundcolor=gray!25,
    linewidth=0pt,
    skipbelow=\topskip,
    skipabove=\topskip
}

\newenvironment{MyShadequote}[1][]{%
    \ignorespaces%
    \begin{mdframed}[style=MyShadeQuoteStyle,#1]%
}{%
    \end{mdframed}%
    \ignorespacesafterend%
}%
% -------- \Quoting ------------

@@

@@

Electronic music is usually defined as music that produced using electronic devices (TODO cite).
In addition to creation techniques, electronic music presents several unique aesthetic characteristics.
These where explored by researchers and composers since the early forays into the field in the late 1940\textsubscript{s}.

@@

Gy{\"o}rgy Ligeti, one of the most innovative musicians of modern 20\textsuperscript{th}-century, was born in Hungary to a Jewish family in 1923.
After suffering from two tyrannies in his youth, Nazi and Stalinist, he left Hungary in 1956 facing western Europe (\cite{ligeti_grove}).
There, he was exposed to electronic music through the Darmstadt-Cologne avant guard ideology, which influenced his writing ever since (\cite[p. TODO]{levy2006}).

@@

The three composers that the current work is about --- Ligeti, Stockhausen and Koenig --- found themselves, each one in its own way, taking part in the Darmstadt-Cologne scene during the 1950\textsuperscript{s}.
Moreover, they all were in a similar stage of their career when they exposed to the ideas presented in the summer courses and the techniques used in the electronic music studios.
These ideas deeply influenced their styles as composers (TODO cite).
However, whereas Stockhausen and Koenig musical styles are direct continuations of these ideas, Ligeti's music cannot be perceived as part of Darmstadt-Cologne style (TODO cite);
his main repertoire is acoustic, and his works are usually not categorized as serial music.

@@

Indeed, researchers often describe Ligeti's music as pale and nervous dream-like (or even nightmarish) reflection of life.
``Grand emotions are captured in a manner that they do not sound real or believable'' argues Frigyesi, ``gestures exaggerate to the point that it becomes surreal, frightening, absurd, or even comical'' (TODO Frigeyesi).

@@

In addition, researchers pointed out that Ligeti's machines always brake down in his music, what makes them only more frightening (TODO cite).

@@

Based on the above, Stockhausen approaches electronic music as a tool for total serializm (TODO total serializm reference).
Furthermore, he argues that serializm must exist in electronic composition, inherently.

@@

This paper deals mainly with serializm, and how it is applied to electronic music.
The following idea came from this way of thinking, in serias, but I believe that it might by understood and analyzed even without considering serial decisions.
Here, Koenig describe different ways to represent a pitch range.

@@

\begin{figure}[!htb]
  \centering
  \textbf{Categorization of ideas}\par\medskip
  \includegraphics[width=\linewidth]{graphics/concepts_categorization.pdf}
  \caption{Subjective categorization of ideas to sound, serializm and high level expressions.}
  \label{fig:concepts_categorization}
\end{figure}

@@

The second category I found is serializm.
As mentioned earlier, I will not deal directly with serializm in the current work.
However, there are some concepts that are rooted in serial way of thinking that can be generalized and understood by themselves.
Concepts like these, despite their problematic source, are not ignored and will be analyzed in the next section.
This category is probably the most foreign one to Ligeti's way of thinking.
Moreover, it seems that a unified model of time, as described by Stockhausen, contradict much of Ligetis' concepts from the next category.

@@

In this section I analyze Ligeti's Ramifications according to the concepts described before.
The order of the concepts in the analysis will follow the three categories described in the last section (\ref{sub:comparison}), starting from sound, through serializm and finally to higher level expressions.
In each category I analyze general concepts first when applicable.

@@

Gesticulated phrases appears in Ramifications only for short periods, they are not characterize the composition as a whole.
Nevertheless, I can note two ``overly exaggerated'' phrases, where the primer is a bit more distinct for that mater: 3:36 - 3:45, 6:28 - 6:40.
One could argue that these are just dramatic phrases within the composition, but I suspect that this amount of vibrato in Ligeti's composition most represent some decent amount of sarcasm (TODO cite frigyesi).

@@

Throughout this work I refrained from analyzing serial concepts per-se, and tried to extract ideas that can stand by themselves from the context of serializm when possible.
As I see it, the fact that Ligeti's music have nothing to do with serializm should not constrain the analysis of its similarity to electronically created music.
Moreover, I think that the analysis of Ramification using both electronic and instrumental concepts support this argument.

Two of Stockhausen arguments about electronic music are hard to sattle and generalize to a `way of thinking' thing.
The first is the statement that electronic music most present serializm, inherently.
A further analysis of Stockhausen's electronic music based on the concepts above may reveal more insights about this issue.
The other one, which is based on the former, is the concept of unified time model.
As already mentioned, it seems that this idea contradicts much of Ligeti's way of thinking.
Again, analysis of Stockhausen's compositions may shed more light about how this concepts are realized, or not, in electronic pieces.
Compared to Stockhausen, Koenigs' concepts apparently works great with Ligeti's music.

>>>>

\mdfdefinestyle{MyShadeQuoteStyle}{
    leftmargin=15pt,
    rightmargin=15pt,
    backgroundcolor=gray!15,
    linewidth=0pt,
}

\newenvironment{MyShadequote}[1][]
    {\begin{mdframed}[style=MyShadeQuoteStyle,#1]}
    {\end{mdframed}}
% -------- \Quoting ------------

@@

% Prevent text from go over the right margin.
% http://tex.stackexchange.com/a/9110/42891
\sloppy

@@

According to the Oxford English Dictionary, the definition of electronic music is ``music performed using synthesizers and other electronic instruments'' (\cite{oxford_electronic_music}).
In addition to creation techniques, electronic music presents several unique aesthetic characteristics.
These where explored by researchers and composers since the early forays into the field in the late 1940\textsubscript{s}.

@@

Gy{\"o}rgy Ligeti, one of the most innovative musicians of modern 20\textsuperscript{th}-century, was born in Hungary to a Jewish family in 1923.
After suffering from two tyrannies in his youth, Nazi and Stalinist, he left Hungary in 1956 facing western Europe (\cite{ligeti_grove}).
There, he was exposed to electronic music through the Darmstadt-Cologne avant guard ideology, which influenced his writing ever since (\cite{levy2006}).

@@

The three composers that the current work is about --- Ligeti, Stockhausen and Koenig --- found themselves, each one in its own way, taking part in the Darmstadt-Cologne scene during the 1950\textsuperscript{s}.
Moreover, they all were in a similar stage of their career when they exposed to the ideas presented in the summer courses and the techniques used in the electronic music studios.
These ideas deeply influenced their styles as composers.
However, whereas Stockhausen and Koenig musical styles are direct continuations of these ideas, Ligeti's music cannot be perceived as part of Darmstadt-Cologne style (\cite{levy2006});
his main repertoire is acoustic, and his works are usually not categorized as serial music.

@@

Indeed, researchers often describe Ligeti's music as pale and nervous dream-like (or even nightmarish) reflection of life.
``Grand emotions are captured in a manner that they do not sound real or believable'' argues Frigyesi, ``gestures exaggerate to the point that it becomes surreal, frightening, absurd, or even comical'' (\cite*{frigyesi}).

@@

In addition, researchers pointed out that Ligeti's machines always brake down in his music, what makes them only more frightening (\cite{frigyesi}).

@@

Based on the above, Stockhausen approaches electronic music as a tool for serialism.
Furthermore, he argues that serialism must exist in electronic composition, inherently.

@@

This paper deals mainly with serialism, and how it is applied to electronic music.
The following idea came from this way of thinking, in serias, but I believe that it might by understood and analyzed even without considering serial decisions.
Here, Koenig describe different ways to represent a pitch range.

@@

\begin{figure}[!htb]
  \centering
  \textbf{Categorization of ideas}\par\medskip
  \includegraphics[width=\linewidth]{graphics/concepts_categorization.pdf}
  \caption{Subjective categorization of ideas to sound, serialism and high level expressions.}
  \label{fig:concepts_categorization}
\end{figure}

@@

The second category I found is serialism.
As mentioned earlier, I will not deal directly with serialism in the current work.
However, there are some concepts that are rooted in serial way of thinking that can be generalized and understood by themselves.
Concepts like these, despite their problematic source, are not ignored and will be analyzed in the next section.
This category is probably the most foreign one to Ligeti's way of thinking.
Moreover, it seems that a unified model of time, as described by Stockhausen, contradict much of Ligetis' concepts from the next category.

@@

In this section I analyze Ligeti's Ramifications according to the concepts described before.
The order of the concepts in the analysis will follow the three categories described in the last section (\ref{sub:comparison}), starting from sound, through serialism and finally to higher level expressions.
In each category I analyze general concepts first when applicable.

@@

Gesticulated phrases appears in Ramifications only for short periods, they are not characterize the composition as a whole.
Nevertheless, I can note two ``overly exaggerated'' phrases, where the primer is a bit more distinct for that mater: 3:36 - 3:45, 6:28 - 6:40.
One could argue that these are just dramatic phrases within the composition, but I suspect that this amount of vibrato in Ligeti's composition most represent some decent amount of sarcasm.

@@

Throughout this work I refrained from analyzing serial concepts per-se, and tried to extract ideas that can stand by themselves from the context of serialism when possible.
As I see it, the fact that Ligeti's music have nothing to do with serialism should not constrain the analysis of its similarity to electronically created music.
Moreover, I think that the analysis of Ramification using both electronic and instrumental concepts support this argument.

Two of Stockhausen arguments about electronic music are hard to sattle and generalize to a `way of thinking' thing.
The first is the statement that electronic music most present serialism, inherently.
A further analysis of Stockhausen's electronic music based on the concepts above may reveal more insights about this issue.
The other one, which is based on the former, is the concept of unified time model.
As already mentioned, it seems that this idea contradicts much of Ligeti's way of thinking.
Again, analysis of Stockhausen's compositions may shed more light about how this concepts are realized, or not, in electronic pieces.
Compared to Stockhausen, Koenigs' concepts apparently works great with Ligeti's music.
